How topos theory can help commutative algebra

### 2018-07-04 slides-como2018.tex
More notes

\title{Exploring the internal language of toposes}
\author{Ingo Blechschmidt}
\date{June 22nd, 2018 XXX}

@@

  \hil{How topos theory can help algebra} \\

@@

\begin{frame}{The internal language of~$\boldsymbol{\Sh(X)}$}
  \small
  Let~$X$ be a topological space. We recursively define
  \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
  \mbox{for open subsets~$U \subseteq X$ and formulas~$\varphi$.
  Write~``$\Sh(X) \models \varphi$'' to mean~$X \models \varphi$.}
  \[ \renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\ }c@{\ }l@{}}
  U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
  U \models \varphi \wedge \psi &\Ll&
  \text{$U \models \varphi$ and $U \models \psi$} \\
  U \models \varphi \vee \psi &\Ll&
  \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $U = \bigcup_i U_i$ s.\,th.} \\
  && \quad\quad\text{for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
  U \models \varphi \Rightarrow \psi &\Ll&
  \text{for all open~$V \subseteq U$: }
  \text{$V \models \varphi$ implies $V \models \psi$} \\
  U \models \forall s \? F\_ \varphi(s) &\Ll&
  \text{for all open $V \subseteq U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
  U \models \forall F\_ \varphi(F) &\Ll&
  \text{for all open $V \subseteq U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
  U \models \exists s \? F\_ \varphi(s) &\Ll&
  \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
  && \quad\quad \text{there exists~$s_0 \in F(U_i)$ s.\,th. $U_i \models \varphi(s_0)$} \\
  U \models \exists F\_ \varphi(F) &\Ll&
  \text{there exists a covering $U = \bigcup_i U_i$ s.\,th. for all~$i$:} \\
  && \quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ s.\,th. $U_i \models \varphi(F_0)$}
  \end{array} \]

@@

  Here is an explicit example of the translation procedure. Let~$\alpha : \F
  \to \G$ be a morphism of sheaves on~$X$. Then:
  \begin{align*}
    & X \models \speak{$\alpha$ is injective} \\[0.5em]
    \Longleftrightarrow\
    & X \models \forall s,t\?\F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
    &\qquad\qquad
        U \models \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.5em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
    &\qquad\qquad
        \text{for all open~$V \subseteq U$:} \\
    &\qquad\qquad\qquad\qquad
          \text{$\alpha_V(s|_V) = \alpha_V(t|_V)$ implies $s|_V = t|_V$} \\[0.5em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s, t \in \F(U)$:} \\
    &\qquad\qquad
          \text{$\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.5em]
    \Longleftrightarrow\ &
      \text{$\alpha$ is a monomorphism of sheaves}
  \end{align*}}
\end{frame}

@@

  Constructively, there are several non-equivalent forms of the intermediate
  value theorem. The following version doesn't admit an intuitionistic proof:
  \begin{indentblock}Let~$g : \RR \to \RR$ be a function between the (Dedekind)
  reals which is continuous in the usual epsilon/delta sense. Assume~$g(-1) < 0
  < g(1)$. Then there exists a number~$x \in \RR$ such that~$g(x) = 0$.
  \end{indentblock}
  If there was an intuitionistic proof, the statement would hold in any topos,
  so in particular in sheaf toposes over topological spaces. By the
  translations shown on the slide (the corner quotes ``$\speak{\ldots}$''
  indicate that translation into formal language is left to the reader), this
  would amount to the following strengthening of the intermediate value
  theorem: In continuous families of continuous functions, zeros can locally be
  picked continuously. However, this strengthening is invalid, as
  \fixedhref{https://raw.githubusercontent.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
  video shows}.

@@

  \jnote{1-3}{The generic prime filter~$\fff_0$ can be described in explicit
  terms. For ring elements~$f$ and~$s$, $D(f) \models (s \in \fff_0)$ iff~$f
  \in \sqrt{(s)}$.}

  \jnote{2-3}{Our description of~$M^\sim$ reveals a precise sense in
  which~$M^\sim$ and~$M$ are related: $M^\sim$ is simply a localization of~$M$
  (first lifted to another universe by the constant sheaf construction). The
  classical descriptions don't make the relation evident.}

  \jnote{3}{One can show, assuming that the little Zariski topos is
  \emph{overt}, that the module~$M$ in~$\Set$ and the module~$\ull{M}$ of the
  little Zariski topos share all first-order properties. This observation
  explains the metatheorem displayed at the bottom left. The assumption is
  satisfied if any element of~$A$ is nilpotent or not nilpotent, so it's always
  satisfied if the law of excluded middle is available. In an intuitionistic
  context, it's still ``morally satisfied''. Details are in Section~12.9 of
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
  notes}.}

  \jnote{3}{As a first approximation, the module~$M^\sim$ can be thought of as
  a reification of all the stalks of~$M$ as a single object. The metatheorem
  makes this precise and also shows the limits of this view: It is only correct
  for geometric sequents. When considering nongeometric sequents, phenomena
  appear which are unique to~$M^\sim$ in the sense that they are in general not
  shared by~$M$, its stalks or its quotients.}

@@

  \jnote{1}{XXX ... succeeded ...}

>>>>

\title{How topos theory can help commutative algebra}
\author{Ingo Blechschmidt}
\date{June 29th, 2018}

@@

  \hil{How topos theory can help commutative algebra} \\

@@

\begin{frame}{The Kripke--Joyal semantics of~$\boldsymbol{\Sh(X)}$}
  \small\vspace*{-0.5em}
  Let~$X$ be a topological space. We recursively define
  \[ U \models \varphi \quad \text{(``$\varphi$ holds on~$U$'')} \]
  \mbox{for open subsets~$U \subseteq X$ and formulas~$\varphi$.
  Write~``$\Sh(X) \models \varphi$'' to mean~$X \models \varphi$.}
  \footnotesize
  \[ \renewcommand{\arraystretch}{1.08}\begin{array}{@{}l@{\ }c@{\ }l@{}}
  U \models \top &\Ll& \text{true} \\
  U \models \bot &\Ll& \hcancel{\text{false}}{0pt}{3pt}{0pt}{-2pt}\ U = \emptyset \\
  U \models s = t \? F &\Ll& s|_U = t|_U \in F(U) \\
  U \models \varphi \wedge \psi &\Ll&
  \text{$U \models \varphi$ and $U \models \psi$} \\
  U \models \varphi \vee \psi &\Ll&
  \hcancel{\text{$U \models \varphi$ or $U \models \psi$}}{0pt}{3pt}{0pt}{-2pt}\ \text{there exists a covering $U = \bigcup_i U_i$} \\
  && \quad\quad\text{such that for all~$i$: $U_i \models \varphi$ or $U_i \models \psi$} \\
  U \models \varphi \Rightarrow \psi &\Ll&
  \text{for all open~$V \subseteq U$: }
  \text{$V \models \varphi$ implies $V \models \psi$} \\
  U \models \forall s \? F\_ \varphi(s) &\Ll&
  \text{for all open $V \subseteq U$ and sections~$s_0 \in F(V)$: $V \models \varphi(s_0)$} \\
  U \models \forall F\_ \varphi(F) &\Ll&
  \text{for all open $V \subseteq U$ and sheaves~$F_0$ over~$V$: $V \models \varphi(F_0)$} \\
  U \models \exists s \? F\_ \varphi(s) &\Ll&
  \hcancel{\text{there exists $s_0 \in F(U)$ such that $U \models \varphi(s_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
  && \quad\quad
  \text{there exists a covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
  && \quad\quad\quad\quad \text{there exists~$s_0 \in F(U_i)$ such that $U_i \models \varphi(s_0)$} \\
  U \models \exists F\_ \varphi(F) &\Ll&
  \hcancel{\text{there exists a sheaf $F_0$ on $U$ such that $U \models
  \varphi(F_0)$}}{0pt}{3pt}{0pt}{-2pt} \\
  && \quad\quad
  \text{there exists a covering $U = \bigcup_i U_i$ such that for all~$i$:} \\
  && \quad\quad\quad\quad \text{there exists a sheaf~$F_0$ on $U_i$ such that $U_i \models \varphi(F_0)$}
  \end{array} \]

@@

  Here is an explicit example of the translation procedure. Let~$\alpha : F
  \to G$ be a morphism of sheaves on~$X$. Then (the corner quotes ``$\speak{\ldots}$''
  indicate that translation into formal language is left to the reader):
  \footnotesize
  \begin{align*}
    & X \models \speak{$\alpha$ is injective} \\[0.2em]
    \Longleftrightarrow\
    & X \models \forall s\?F\_ \forall t\?F\_ \alpha(s) = \alpha(t) \Rightarrow s = t \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
    &\qquad\qquad
        U \models \forall t\?F\_ \alpha(s_0) = \alpha(t) \Rightarrow s_0 = t \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
    &\qquad\qquad
        \text{for all open~$V \subseteq U$, sections $t_0 \in F(V)$:} \\
    &\qquad\qquad\qquad\qquad
          V \models \alpha(s_0) = \alpha(t_0) \Rightarrow s_0 = t_0 \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s_0 \in F(U)$:} \\
    &\qquad\qquad
        \text{for all open~$V \subseteq U$, sections $t_0 \in F(V)$:} \\
    &\qquad\qquad\qquad\qquad
          \text{for all open $W \subseteq V$: $\alpha_V(s_0|_W) = \alpha_V(t_0|_W)$ implies $s_0|_W = t_0|_W$} \\[0.2em]
    \Longleftrightarrow\ &
      \text{for all open~$U \subseteq X$, sections $s, t \in F(U)$:
        $\alpha_U(s|_U) = \alpha_U(t|_U)$ implies $s|_U = t|_U$} \\[0.2em]
    \Longleftrightarrow\ &
      \text{$\alpha$ is a monomorphism of sheaves}
  \end{align*}}
\end{frame}

@@

  Constructively, there are several non-equivalent forms of the intermediate
  value theorem. The following version doesn't admit an intuitionistic proof:
  \begin{indentblock}Let~$g : \RR \to \RR$ be a function between the (Dedekind)
  reals which is continuous in the usual epsilon/delta sense. Assume~$g(-1) < 0
  < g(1)$. Then there exists a number~$x \in \RR$ such that~$g(x) = 0$.
  \end{indentblock}
  If there was an intuitionistic proof, the statement would hold in any topos,
  so in particular in sheaf toposes over topological spaces. By the
  translations shown on the slide, this
  would amount to the following strengthening of the intermediate value
  theorem: In continuous families of continuous functions, zeros can locally be
  picked continuously. However, this strengthening is invalid, as
  \fixedhref{https://raw.githubusercontent.com/iblech/internal-methods/master/images/zeros-in-families.mp4}{this
  video shows}.

@@

  \jnote{1}{The Kripke--Joyal semantics for the little Zariski topos amounts to
  the following: $\Spec(A) \models \varphi$ iff~$D(1) \models \varphi$, and the
  clauses for~$D(f) \models \varphi$, where~$f$ ranges over the elements
  of~$A$, are given by the following table.
  \[\footnotesize\renewcommand{\arraystretch}{1.25}\begin{array}{@{}l@{\quad}c@{\quad}l@{}}
    D(f) \models \top &\text{iff}& \text{true} \\
    D(f) \models \bot &\text{iff}& \text{$f$ is nilpotent} \\
    D(f) \models x = y &\text{iff}& x = y \in A[f^{-1}] \\
    D(f) \models \varphi \wedge \psi &\text{iff}&
      \text{$D(f) \models \varphi$ and $D(f) \models \psi$} \\
    D(f) \models \varphi \vee \psi &\text{iff}&
      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi$ or $D(fg_i) \models \psi$} \\
    D(f) \models \varphi \Rightarrow \psi &\text{iff}&
      \text{for all~$g \in A$, $D(fg) \models \varphi$ implies $D(fg) \models \psi$} \\
    D(f) \models \forall x\?M^\sim\_ \varphi(x) &\text{iff}&
      \text{for all~$g \in A$ and $x_0 \in M[(fg)^{-1}]$, $D(fg) \models \varphi[x_0/x]$} \\
    D(f) \models \exists x\?M^\sim\_ \varphi(x) &\text{iff}&
      \text{there exists a partition~$f^n = fg_1 + \cdots + fg_m$ with,} \\
    &&\quad\text{for each~$i$, $D(fg_i) \models \varphi[x_0/x]$ for some~$x_0 \in M[(fg_i)^{-1}]$}
  \end{array} \]
  The generic prime filter~$\fff_0$ can also be described in explicit
  terms. For ring elements~$f$ and~$s$, $D(f) \models (s \in \fff_0)$ iff~$f
  \in \sqrt{(s)}$.}

  \jnote{2}{Our description of~$M^\sim$ reveals a precise sense in
  which~$M^\sim$ and~$M$ are related: $M^\sim$ is simply a localization of~$M$
  (first lifted to another universe by the constant sheaf construction). The
  classical descriptions don't make the relation evident.

  As a first approximation, the module~$M^\sim$ can be thought of as a
  reification of all the stalks of~$M$ as a single object. The metatheorem
  displayed at the top left on the next slides makes this precise and also
  exposes the limits of this view: It is only correct for geometric sequents.
  When considering nongeometric sequents, phenomena appear which are unique
  to~$M^\sim$ in the sense that they are in general not shared by~$M$, its
  stalks or its quotients.}

  \jnote{3}{One can show, assuming that the little Zariski topos is
  \emph{overt}, that the module~$M$ in~$\Set$ and the module~$\ull{M}$ of the
  little Zariski topos share all first-order properties. This observation
  explains the metatheorem displayed at the bottom left. The assumption is
  satisfied if any element of~$A$ is nilpotent or not nilpotent, so it's always
  satisfied if the law of excluded middle is available in the metatheory. In an
  intuitionistic context, it's still ``morally satisfied''. Details are in
  Section~12.9 of
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
  notes}.

  The metatheorem soups up a number of lemmas of algebraic geometry, there
  stated in geometric language. For instance, if~$M$ is finitely generated,
  then~$M^\sim$ is of finite type. If~$M$ is finitely presented, then~$M^\sim$
  is of finite presentation. If~$M$ is coherent, then~$M^\sim$ is coherent.}

@@

  \jnote{1}{This slide delivers on the promise made earlier: Using the internal
  language of the little Zariski topos, we can reduce to the case of fields
  without having to employ maximal ideals, prime ideals or minimal prime ideals.
  Since the internal language machinery is itself constructive, the displayed
  proofs can be unwinded to yield external constructive proofs which don't
  reference topos theory. Details on these external proofs can soon be found in
  a forthcoming paper titled
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-wlog.pdf}{Without
  loss of generality, any reduced ring is a field}.}

  \jnote{2}{The testcase of Grothendieck's generic freeness lemma illustrates
  that the internal language of toposes can help commutative algebra even if
  one is not interested in constructivity issues. The previously known proofs
  are somewhat long and somewhat convoluted; the new proof is arguably short
  and simple.

  Details on the internal proof are in Section~11.5 of
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
  notes}. The
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-generic-freeness.pdf}{external
  proof} obtained by unwinding the internal one is still quite direct, compared
  to the previously published proofs, and interestingly follows a curious
  course: It starts with verifying, in an inductive manner, that~$B$ and~$M$
  are free; that~$A \to B$ is of finite presentation; and that~$M$ is finitely
  presented as a~$B$-module. Then the assumption for~$f = 1$ is used, rendering
  the prior steps moot, since over the zero ring any module is free and any
  algebra is of finite presentation. External translations of internal proofs
  which use double negation will always take such a course.}

  \jnote{3}{XXX}
